Predicting Customer Behavior Using a Portuguese Financial Institution Dataset 

### 2024-08-05 manuscript/Manuscript.qmd
Last Major Changes

---
title: "Predicting Customer Behavior Using a Portuguese Financial Institution Dataset "
subtitle: ""
author: Antonio Flores
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

@@

Financial institutions spend billions of dollars on their marketing teams every year, with a positive trend observed in 2023 for the largest and smallest banks (@akins2024). With the need for efficient marketing becoming more and more important, machine learning models are viable tools for refining marketing strategies. This project will seek to identify a model for predicting consumer financial behavior using a dataset from a Portuguese bank by comparing the accuracy and predictive power of different machine learning models.

@@

Rows: 45, 211\
Columns: 17 {{< pagebreak >}}

@@

I removed the 'Duration' variable due to a warning I discovered on the site hosting this dataset initially: "Important note: this attribute highly affects the output target (e.g., if duration=0 then y='no'). Yet, the duration is not known before a call is performed. Also, after the end of the call y is obviously known. Thus, this input should only be included for benchmark purposes and should be discarded if the intention is to have a realistic predictive model."(@misc_bank_marketing_222) Due to this new information I decided to remove all the variables related to this, which included 'campaign', 'pdays', 'previous', and 'poutcome.' Bringing our new total number of predictors to 11.

@@

@fig-result2 shows the distribution of the 'Education' variable

@@

@fig-result3 shows the most common 'Job' types in our data

@@

@fig-result4 shows a scatter plot with Age and bank account balance as the (x,y) values, stratified by marital status.

```{r}
#| label: fig-result4
#| fig-cap: "Scatterplot of Age and Balance"
#| echo: FALSE
knitr::include_graphics(here("results","figures","age-balance-stratified.png"))
```

@@

While there doesn't appear to be a significant trend, we can clearly see that the 30th of the month stands out as being a day of interest, especially compared to other days of the months. 
{{< pagebreak >}}

@@

I chose to utilize the following models: the Multivariate Adaptive Regression Splines model (MARS), K-Nearest Neighbors model (KNN), Logistic Regression model, and RandomForest models to determine the best prediction model for this project.

@@

```{r}
#| label: fig-result999
#| fig-cap: "MARS Model example from MiniTab"
#| echo: FALSE
knitr::include_graphics(here("results","images","marsexample.jpeg"))
```
{{< pagebreak >}}

@@

**RandomForest:**\
RandomForest models take advantage of decision trees. If we think about the scenario in our project (whether someone makes a deposit or not), we could imagine a decsion tree starting with, "after what age is someone more likely to subscribe?" This would be our first node to split the data on. We could continue asking things like, "Are those with housing loans more likely to subscribe?" or "Are those who are married more liklely to subscribe?", and these would represent more decision nodes for us to split the data on, getting us closer to the mode accurate prediction model. The RandomForest algorithm uses different methods to create several uncorrelated "forests" of decision trees(@IBM) .
{{< pagebreak >}}
### Background Regarding Performance Metrics

@@

**Recall:**\
The proportion of Positive cases that were correctly identified. This helps us understand how well we can classify positive cases specifically.\

@@

**F1:**\
The harmonic mean of Precision and Recall. This metric helps iron out extreme values\
{{< pagebreak >}}

@@

```{r}
#| label: tbl-resulttable1
#| echo: FALSE
resulttable1 = readRDS(here("results","tables","finaltable1.rds"))
knitr::kable(resulttable1)
```

@@

```{r}
#| label: tbl-resulttable2
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","finaltable2.rds"))
knitr::kable(resulttable2)
```

With these results we can see that the RandomForest Model is still the best performing, but in this case, the KNN model is not too far behind in most metrics. The Logistic Regression model performed better under these conditions, predicting some positive classes. However, the MARS model did not fare so well. In this run, the MARS model was the model to not identify a single positive class.
{{< pagebreak >}}

@@

@fig-result7 shows the Predictors that had the most impact on the Mars Model -- Numeric

```{r}
#| label: fig-result7
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantM1.png"))
```

@@

@fig-result9 shows the Predictors that had the most impact on the KNN Model--Numeric

```{r}
#| label: fig-result9
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantKNN1.png"))
```

@@

@@

  Despite these findings, the models were able to produce which variables provided the most weight towards the prediction power. The Housing variable was listed as the most important variable by every model with the exception of the Mars-Numeric model. Balance and Age were also listed as important by several different models. The variables that did not show up in the five important variables in any of the models were: Marital, Job, and Default. Education, Loan, and Day had a significant effect on some of the models and little to no effect on others.

>>>>

---
title: "Predicting Customer Behavior Using a Portuguese Financial Institution Dataset "
subtitle: ""
author: Antonio Flores
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---

@@

Financial institutions spend billions of dollars on their marketing teams every year, with a positive trend observed in 2023 for the largest and smallest banks (@akins2024). With the need for efficient marketing becoming more and more important, machine learning models are viable tools for refining marketing strategies. This project will seek to identify a model for predicting consumer financial behavior using a dataset from a Portuguese bank. Through comparing the accuracy and predictive power of four different machine learning models, this project will identify an ideal method for prediction.

@@

Rows: 45, 211\
Columns: 17

@@

I removed the 'Duration' variable due to a warning I discovered on the site hosting this dataset initially: "Important note: this attribute highly affects the output target (e.g., if duration=0 then y='no'). Yet, the duration is not known before a call is performed. Also, after the end of the call y is obviously known. Thus, this input should only be included for benchmark purposes and should be discarded if the intention is to have a realistic predictive model."(@misc_bank_marketing_222) Due to this new information I decided to remove all the variables related to this, which included 'campaign', 'pdays', 'previous', and 'poutcome.' Bringing our new total number of predictors to 11.
#{{< pagebreak >}}

@@

@@

@@

```{r}
#| label: fig-result4
#| fig-cap: "Scatterplot of Age and Balance, stratified by Marital Status."
#| echo: FALSE
knitr::include_graphics(here("results","figures","age-balance-stratified.png"))
```

@@

While there doesn't appear to be a significant trend, we can clearly see that the 30th of the month stands out as being a day of interest, especially compared to other days of the months.

@@

I chose to utilize the following models: the Multivariate Adaptive Regression Splines model (MARS), K-Nearest Neighbors model (KNN), Logistic Regression model, and RandomForest models to determine the best prediction model for this project.
{{< pagebreak >}}

@@

```{r}
#| label: fig-result7
#| fig-cap: "MARS Model example from MiniTab"
#| echo: FALSE
knitr::include_graphics(here("results","images","marsexample.jpeg"))
```

@@

**RandomForest:**\
RandomForest models take advantage of decision trees. If we think about the scenario in our project (whether someone makes a deposit or not), we could imagine a decsion tree starting with, "after what age is someone more likely to subscribe?" This would be our first node to split the data on. We could continue asking things like, "Are those with housing loans more likely to subscribe?" or "Are those who are married more liklely to subscribe?", and these would represent more decision nodes for us to split the data on, getting us closer to the mode accurate prediction model. The RandomForest algorithm uses different methods to create several uncorrelated "forests" of decision trees(@IBM) .
{{< pagebreak >}}
### Background On Performance Metrics

@@

**Recall:(Sensitivity) **\
The proportion of Positive cases that were correctly identified. This helps us understand how well we can classify positive cases specifically.\

**Specificity:**\
The proportion of Negative cases that were correctly identified.

@@

**F1:**\
The harmonic mean of Precision and Recall. This metric helps iron out extreme values\

**ROC:**\
The Receiver Operator Characteristic is a plotting technique that shows showing the threshold difference between Sensitivity (Recall) and Specificity. A model with good performance will have curve closer to the top left of the graph, whereas, a low-performing model will have a curve closer to the middle diagonal line.

```{r}
#| label: fig-result8
#| fig-cap: "ROC diagram from DisplayR"
#| echo: FALSE
knitr::include_graphics(here("results","images","ROC_Example.jpg"))
```

@@

```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Performance Metrics for Numeric Unfactored Model Run"
#| echo: FALSE
resulttable1 = readRDS(here("results","tables","finaltable1.rds"))
knitr::kable(resulttable1)
#: My table {#tbl-resulttable1}
```

@@

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Performance Metrics for Factored Model Run"
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","finaltable2.rds"))
knitr::kable(resulttable2)
```

With these results we can see that the RandomForest Model is still the best performing, but in this case, the KNN model is not too far behind in most metrics. The Logistic Regression model performed better under these conditions, predicting some positive classes. However, the MARS model did not fare so well. In this run, the MARS model was the model to not identify a single positive class.

@fig-result9 displays the ROC Curve for the first model run. Both ROC charts were identical, so only one is included here.

```{r}
#| label: fig-result9
#| echo: FALSE
knitr::include_graphics(here("results","figures","ROC1.png"))
```
The ROC graph tells an even clearer story. While the RF model had a near perfect threshold trade-off between Sensitivity and Specificity, the other three models had very low ROC scores, again emphasizing poor predictive performance.

@@

@fig-result10 shows the Predictors that had the most impact on the Mars Model -- Numeric

```{r}
#| label: fig-result10
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantM1.png"))
```

@@

@fig-result11 shows the Predictors that had the most impact on the KNN Model--Numeric

```{r}
#| label: fig-result11
#| echo: FALSE
knitr::include_graphics(here("results","figures","importantKNN1.png"))
```

@@

## DownSampling Machine Learning Models
In situations involving imbalanced response variables, there are different approaches available that can improve model performance. Kuhn and Johnson describe several different options including "Up-sampling" and "Down-sampling" (@kuhn2018). Up-sampling involves generating additional synthetic data in order to bring the totals of the response classes up to an equal amount. Down-sampling involves removing random observations until the the majority class is equal to the minority class (or classes). Since this project was already utilizing a fairly large dataset, I decided to perform Down-sampling in order to assist with the response variable imbalance that was being observed. The challenge with this approach is that there is less data to train with, but could lead to a greater overall model performance.

I performed Down-Sampling on the training data set but kept the test data set intact as it was. It is recommended to allow the test data set to continue representing the natural data distribution in order for the most realistic prediction analysis

### DownSampling Model Performance Metrics (Non-factored)

```{r}
#| label: tbl-resulttable3
#| echo: FALSE
resulttable3 = readRDS(here("results","tables","finaltable3.rds"))
knitr::kable(resulttable3)
```
At first glance, it appears that the performance power on these models is less than the first models, but this is not the case. While Accuracy was lower among all models, Recall and F1 scores saw tremendous improvements. The Model that improved the most was clearly Logistic Regression which didn't show any Precision or F1 power initially, but now shows better Precision and F1 than the RandomForest model, which still showed to the be the superior model after this run.

### DownSampling Model Performance Metrics (Factored)

```{r}
#| label: tbl-resulttable4
#| echo: FALSE
resulttable4 = readRDS(here("results","tables","finaltable4.rds"))
knitr::kable(resulttable4)
```
We can observe similar changes in the second Down-sampled run. In the Factored-Down-Sampled model runs, we also see that it is harder to identify a single model of best performance. The MARS model showed the best, Accuracy and Specificity, while the RandomForest model showed the best Sensitivity and Precision. The Logistic Regression model was not far behind in these categories and managed to approximately tie three ways for the F1 metric, along with the MARS Model and RandomForest Model. 
{{< pagebreak >}}

@tbl-resulttable15 displays the ROC Curve for the Down-Sampled model runs. Again, both ROC charts were identical, so only one is included here.

```{r}
#| label: tbl-resulttable15
#| echo: FALSE
knitr::include_graphics(here("results","figures","ROC3.png"))
```
We can see much better results on the ROC model compared to the first runs. The RandomForest again performs nearly perfectly, but all three remaining models show some promise in terms of threshold trade-off, with the MARS model leading these three.

### Variables of Importance for Down-Sampled Models

@fig-result16 shows the Predictors that had the most impact on the Down-Sampled Mars Model -- Factored

```{r}
#| label: fig-result16
#| echo: FALSE
knitr::include_graphics(here("results","figures","DSimportantM2.png"))
```
Both MARS models highlighted essentially the same variables of importance, with added detail coming from the factored model. As mentioned previously, Housing2 indicates that the person does have a housing loan and Loan2 indicates that the person has personal loan. Education3 indicates that the person has attained a tertiary level of education. Marital 2 indicates the person is married.
{{< pagebreak >}}

@fig-result17 shows the Predictors that had the most impact on the Down-Sampled KNN Models (both models highlighted the same variables)

```{r}
#| label: fig-result17
#| echo: FALSE
knitr::include_graphics(here("results","figures","DSimportantKNN1.png"))
```
The variables highlighted here are the same that were highlighted in the original KNN models with the only exception being the Marital variable moved up one spot on this list.
{{< pagebreak >}}

@fig-result18 shows the Predictors that had the most impact on the Down-Sampled Logistic Regression Model -- Factored

```{r}
#| label: fig-result18
#| echo: FALSE
knitr::include_graphics(here("results","figures","DSimportantLog2.png"))
```
These are practically the same variables highlighted in the original Logistic Regression Model, except while the 'Education3' variable was the third most important variable on the original model, it is nowhere to be found on this new list. In contrast, the 'Job3' (Technician) variable makes an appearance as having a substantial impact on prediction power.
{{< pagebreak >}}

@fig-result19 shows the Predictors that had the most impact on the Down-Sampled Logistic Regression Model -- Factored

```{r}
#| label: fig-result19
#| echo: FALSE
knitr::include_graphics(here("results","figures","DSimportantRF1.png"))
```
While the two Down-Sampled models were fairly close in highlighting variables, there was quite a bit of difference between these results and the original RandomForest model. Where the original highlighted, Age, Housing, Day and Balance, the new models highlighted Housing, Age, Balance, and Loan.
{{< pagebreak >}}

@@

  Using a Down-Sampling approach to alleviate the imbalance issue resulted in models with greater predictive power and gave more clarity on which variables (and in some cases, sub-variables) contributed the most towards correctly predicting the response variable.

  All four models provided insights for this objective. The Housing variable was listed as the most important variable by every model with the exception of the Mars-Numeric model. Balance and Age were also listed as important by several different models. The variables that did not show up in the five important variables in any of the models were: Marital, Job, and Default. Education, Loan, and Day had a significant effect on some of the models and little to no effect on others.
